Applying Radial Basis Function Networks and Markov Chains for on-line detection of concept drift in non-stationary environments

### 2019-10-30 tex/article/article.tex
cd review

Most real-world problems scenarios can be regarded as non-stationary environments \cite{Gama:2014:DAF:2670967.2670971}.
In these environments, the joint probability distribution can change over time,
such as a switch in the conditional probability distribution on a classification problem,
or a change of some moment (such as mean and variance) on a time series forecasting problem \cite{tsymbal2004problem}.
%
Systems applied to these settings may be unable to adapt to the changes, hence dramatically deteriorating their performance.
This phenomenon is known as concept drift.

The Bayesian Theory can be used to formally define the concept drift phenomenon \cite{Elwell:2011}:
consider the posterior probability of a sample $x$ belonging to a class $y$, a concept drift happens when this probability changes over time, that is, $Pt + 1 (y | x) \neq Pt (y | x)$. In a supervised learning scenario, this can be interpreted as when the relationship between the input data and the target variable change over time.

@@

Concept drift detection methods characterize and quantify concept drifts through the delimitation of moments or time intervals in which change happens \cite{Basseville:1993:DAC:151741}.
%
These methods fall into two categories, according to the need for data labeling \cite{Zliobaite:2010}.
The categories are described below and summarized in Table \ref{tbl:algorithms}, together with the main algorithms found in the literature.

\begin{itemize}
    \item \textbf{Explicit Algorithms/Supervised}: Adopt a reactive approach, as they depend on the correct labeling of the data.
    The model performance is monitored continuously, and drifts are detected when its performance starts to deteriorate, passing past a threshold.
    \item \textbf{Implicit Algorithms/Unsupervised}: Implement a proactive approach and are independent of data labels.
    Concept drifts are detected through the analysis of incoming data or indicators produced by the applied learning techniques.
    Although more prone to false alarms, they are an alternative to scenarios where obtaining labels is expensive, time-consuming, or unviable.
    Also, this approach can lead to better results, since it is possible to refit the model or adjust the data before the deterioration of the predictions.
\end{itemize}

@@

The method proposed in this paper classifies as an \textbf{Implicit/Unsupervised} algorithm and is suited to detect concept drifts in any pattern.
It continually groups the incoming data through Radial Basis Function Networks and maintains a model of the center transitions in a Markov Chain.
Concept drifts are detected when the probability of a transition in the formed cluster reaches a parametric threshold.
This design allows the method to detect concept drifts in real-time and independently of labels.

>>>>

Most real-world problems can be regarded as non-stationary environments \cite{Gama:2014:DAF:2670967.2670971}.
In these scenarios, the joint probability distribution can change over time,
such as a switch in the conditional probability distribution on a classification problem,
or a change of some moment (such as mean and variance) on a time series forecasting problem \cite{tsymbal2004problem}.
%
Systems applied to these settings may be unable to adapt to the changes, hence dramatically deteriorating their performance.
This phenomenon is known as concept drift.

The concept drift phenomenon can be formally defined using Bayesian Theory \cite{Elwell:2011}:
consider the posterior probability of a sample $x$ belonging to a class $y$, a concept drift happens when this probability changes over time, that is, $Pt + 1 (y | x) \neq Pt (y | x)$. In a supervised learning scenario, this can be interpreted as when the relationship between the input data and the target variable change over time.

@@

Concept drift detection methods characterize and quantify concept drifts through the delimitation of moments or time intervals in which change happens \cite{Basseville:1993:DAC:151741}.
%
These methods fall into two categories \cite{Zliobaite:2010},
which are described below and listed in Table \ref{tbl:algorithms}, along with their main algorithms according to the literature.

\begin{itemize}
    \item \textbf{Explicit Algorithms/Supervised}: Adopt a reactive approach, as they depend on the correct labeling of the data.
    The model performance is monitored continuously and concept drifts are detected when it starts to deteriorate, reaching a threshold.
    \item \textbf{Implicit Algorithms/Unsupervised}: Implement a proactive approach and are independent of data labels.
    Concept drifts are detected through the analysis of incoming data or indicators produced by the applied learning techniques.
    Although more prone to false alarms, they are an alternative to scenarios where obtaining labels is expensive, time-consuming, or unviable.
    Also, this approach can lead to better results, since it is possible to refit the model or adjust the data before the deterioration of the predictions.
\end{itemize}

@@

The method proposed in this paper classifies as an \textit{Implicit/Unsupervised} algorithm.
The technique continually groups the incoming data through Radial Basis Function Networks and maintains a model of the center transitions in a Markov Chain.
Concept drifts are detected when the probability of a transition in the formed cluster reaches a parametric threshold.
This design allows the method to detect concept drifts in real-time and independently of labels.
